---
title: "A CS-SEM analysis workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A CS-SEM analysis workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(cssem)

item5 <- function(x) pmin(pmax(as.integer(round(x) + 3L), 1L), 5L)
make_survey_data <- function(n, seed, add_missing = FALSE) {
  set.seed(seed)
  trust_state <- stats::rnorm(n)
  age <- stats::rnorm(n, mean = 40, sd = 8)
  loyalty_state <- .45 * trust_state + .02 * age + stats::rnorm(n)
  data <- data.frame(
    trust_1 = item5(trust_state + stats::rnorm(n, sd = .6)),
    trust_2 = item5(trust_state + stats::rnorm(n, sd = .7)),
    trust_time = trust_state + stats::rnorm(n, sd = .5),
    loyalty_1 = item5(loyalty_state + stats::rnorm(n, sd = .6)),
    loyalty_2 = item5(loyalty_state + stats::rnorm(n, sd = .7)),
    age = age
  )
  if (add_missing) data$trust_2[seq(5L, n, by = 19L)] <- NA_integer_
  data
}
```

This vignette walks through a small, deterministic study from scale declarations
to prediction and support checks. It uses simulated records so it can run without
network access. The sample size and resampling counts are deliberately modest;
use the study's full design and stronger convergence settings for substantive
analysis.

## Declare scales and inspect sample coverage

`mixed_items()` combines existing ordinal and continuous item declarations; it
does not infer scales from the observed values. `partial` missingness lets each
observed indicator contribute to the respondent's measurement posterior.

```{r fit, warning=FALSE}
survey_data <- make_survey_data(120L, 2026L, add_missing = TRUE)
new_data <- make_survey_data(30L, 2027L)
model <- specify_measurement(
  Trust = mixed_items(ordinal("trust_1", "trust_2"), continuous("trust_time")),
  Loyalty = ordinal("loyalty_1", "loyalty_2"),
  Age = manifest("age"), folds = 3
)
fit <- fit_states(model, survey_data, seed = 23, iterations = 6,
  diagnostics = FALSE, missing_policy = "partial", retain_data = FALSE)
sample_accounting(fit)
```

The fit keeps locked construct scores and encoders but omits the training data
frames. `sample_accounting()` preserves row positions and counts. Functions that
need original item rows, such as measurement assessment and bootstrap refits,
require a fit made with `retain_data = TRUE`. Increase `iterations` when the
measurement engine reports that a construct reached its iteration cap; a
suppressed warning in this compact example does not imply convergence.

The provenance record keeps the call, resolved options, software versions,
schema, row positions, and split metadata without embedding observations:

```{r provenance}
cssem_provenance(fit)
```

## Fit and inspect the associational structure

Construct states are standardized scores. The declared paths below are
associational predictive relationships, not automatically causal effects.

```{r association}
structure <- specify_structure(
  Loyalty ~ linear(Trust) + linear(Age),
  order = c("Age", "Trust", "Loyalty")
)
association <- associate(fit, structure, structural_repeats = 1,
  shadow_scope = "temporal", seed = 24)
effect_ledger(association)
cssem_provenance(association)
```

## Contrasts and new-record prediction

The marginal contrast compares the expected outcome at two locked-score values.
The interval is conditional on the selected structural shape. Prediction scores
the new records with the fitted measurement encoders and reports the structural
outcome on the locked-score scale.

```{r prediction}
contrast <- marginal_contrast(association, "Loyalty", "Trust", values = c(-1, 1),
  reps = 40, seed = 27)
contrast
cssem_provenance(contrast)
predict(association, new_data, outcomes = "Loyalty")
```

## Check held-out performance

Outer validation refits measurement and structural selection on each training
partition. Its `outer_test` metrics refer to held-out records; internal selector
metrics remain separately labeled and must not be presented as held-out results.

```{r outer-validation}
splits <- make_splits(survey_data, method = "random", folds = 3, seed = 25)
outer <- validate_outer(model, structure, survey_data, splits, seed = 26,
  iterations = 6, diagnostics = FALSE,
  structural_args = list(structural_repeats = 1, shadow_scope = "temporal"))
outer
```

## Optional causal estimand

This call demonstrates the package's declared adjusted-linear estimand. Its
interpretation requires a defensible temporal order, an adequate adjustment set,
consistency, positivity, and no unmeasured confounding. The software checks some
structural and temporal-order constraints and can audit a typed causal graph,
but it cannot verify those study assumptions from the data or from the
declarations. The result label is not evidence that the assumptions hold.

```{r causal}
trust_loyalty_design <- causal_design(
  edges = data.frame(
    from = c("Age", "Age", "Trust"),
    to = c("Trust", "Loyalty", "Loyalty")
  ),
  treatment = "Trust", outcome = "Loyalty", adjust = "Age",
  assumptions = list(
    consistency = "assumed",
    no_unmeasured_confounding = "assumed",
    positivity = "assumed",
    measurement_validity = "assumed",
    temporal_order = "assumed"
  )
)
causal_effect(association, treatment = "Trust", outcome = "Loyalty",
  adjust = "Age", estimand = "adjusted_linear", disattenuate = FALSE,
  temporal_order = c("Age", "Trust", "Loyalty"),
  design = trust_loyalty_design)
```

The [causal validation manifest](../R/causal-validation.R) also provides
independent structural targets for six score-level stress scenarios; it does
not refit measurement models or establish identification for the example data.

For the estimator's current evidence and explicit boundaries, see
[`capabilities.md`](../docs/capabilities.md) and the
[`method specification`](../docs/method-spec.md).
