---
title: "Study-specific simulation and sample-size planning"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Study-specific simulation and sample-size planning}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(cssem)
```

The study-simulation workflow takes a scenario grid and three callbacks: one
generates the observed data, one returns an independently derived target, and
one runs the analysis. The target callback receives the sample size, so it can
represent quantities that change with the design. It is evaluated before the
replications and is never passed to the analysis callback.

This small Gaussian example estimates a population mean and the expected
prediction error from predicting a new response with the sample mean. If the
response variance is `sigma2`, the expected prediction error is
`sigma2 * (1 + 1 / n)`: one variance term comes from the new response and the
other from estimating the mean with `n` observations.

```{r define-study}
scenarios <- data.frame(
  scenario = "normal_intercept",
  mu = 0.5,
  sigma2 = 0.5
)

generate <- function(scenario, n, seed) {
  set.seed(seed)
  data.frame(y = stats::rnorm(n, mean = scenario$mu, sd = sqrt(scenario$sigma2)))
}

truth <- function(scenario, n) {
  c(mean = scenario$mu,
    prediction_risk = scenario$sigma2 * (1 + 1 / n))
}

analyze <- function(data, scenario, seed) {
  n <- nrow(data)
  estimate_mean <- mean(data$y)
  sample_variance <- stats::var(data$y)
  df <- n - 1L
  mean_half_width <- stats::qt(.975, df = df) * sqrt(sample_variance / n)
  variance_interval <- c(
    df * sample_variance / stats::qchisq(.975, df = df),
    df * sample_variance / stats::qchisq(.025, df = df)
  )
  risk_multiplier <- 1 + 1 / n
  estimates <- data.frame(
    estimand = c("mean", "prediction_risk"),
    estimate = c(estimate_mean, sample_variance * risk_multiplier),
    estimate_status = "available",
    lower = c(estimate_mean - mean_half_width,
      variance_interval[[1L]] * risk_multiplier),
    upper = c(estimate_mean + mean_half_width,
      variance_interval[[2L]] * risk_multiplier),
    interval_status = "available",
    status_reason = "",
    stringsAsFactors = FALSE
  )
  list(status = "completed", converged = TRUE, failure_reason = "",
    estimates = estimates)
}

spec <- study_spec(
  scenarios = scenarios,
  sample_sizes = c(40L, 80L),
  generate = generate,
  truth = truth,
  analyze = analyze,
  metadata = list(
    description = "Gaussian intercept-only prediction study",
    analysis_scope = "refit the sample mean and Gaussian variance at each replication",
    target = "Expected new-response squared prediction error"
  )
)
```

Run replications and summarize them with an explicit null for mean detection:

```{r run-study}
simulation <- simulate_study(spec, reps = 200L, seed = 2026L)
summary <- summarize_study(simulation, null_values = c(mean = 0))

unique(simulation$replications[
  simulation$replications$estimand == "prediction_risk",
  c("n", "truth")
])
```

The table shows a different prediction-risk target at each sample size. The
summary reports both conditional rates among available results and unconditional
rates over every scheduled replication. Failed analyses and unavailable
intervals count as uncovered and as non-detections unconditionally; they remain
visible in the replication ledger and in the failure and availability metrics.

```{r coverage-summary}
subset(summary$metrics,
  metric == "coverage" & estimand == "prediction_risk",
  select = c(scenario, n, estimand, scope, estimate, denominator,
    mcse, mc_lower, mc_upper, availability_status)
)
```

This example validates the callback workflow against a known Gaussian target;
it does not validate the package's CS-SEM estimators. A CS-SEM study can call its
`fit_states()`, `associate()`, and inference pipeline from `analyze()`, while
`generate()` and `truth()` encode the study's measurement design and independent
data-generating target. The declared `analysis_scope` describes what the
callback refits, but it does not prove that every intended pipeline stage is
rerun.

The sample-size planner remains gated on the G4 and G6 methodological evidence
recorded in `docs/gaps.md`; a simulation summary alone is not a universal sample
size rule.
